### 2024-03-26 doc/sections/mav-semantics.tex
[ tex ] done most of the semantics section

\section{Semantics}\label{sec:mav-semantics}

@@

\subsection{$*$-autonomous and Duoidal Structure on Partial Orders}

@@

@@

\begin{remark}
  \bob{A frame is nearly two duoidal structures, except that we don't
    assume that $+$ is a monoid, nor is there a unit element for
    it. If we assumed that $+$ was a lattice join, then (iii) and (iv)
    would automatically be true. We do not include this because these
    facts are not true for the cut-free MAV system.}
\end{remark}

\begin{remark}
  \bob{Connection to process algebras: we can read the ordering as
    ``reduces to'' and the connectives as parallel combination,
    sequential composition, and choice, respectively. If we add in
    atomic elements with duals and an interaction law, then we get
    something like CCS without restriction, but with sequential
    composition.}
\end{remark}

@@

\bob{Some introductory text here: explain the steps in the construction and what each one adds.}

\subsubsection{Lower Sets and Day pomonoids}

@@

\begin{proposition}
  For any $(A, \leq)$, the function $\eta$ is monotone, and
  $(\LowerSet{A}, \subseteq)$ has meets and joins given by
  intersection and union respectively.

  \bob{and $\eta$ preserves any meets that exist}
\end{proposition}

\begin{proposition}\label{prop:day-construction}
  If $(\bullet, i)$ is a pomonoid on $(A, \leq)$, then there is a
  corresponding \emph{Day pomonoid} $(\Day{\bullet}, \Day{i})$ on
  $\LowerSet{A}$ defined as
  $F \Day\bullet G = \{ z \mid z \leq x \bullet y, x \in F, y \in G
    \}$ and $\Day{i} = \eta(i)$. Moreover:
  \begin{enumerate}
    \item $(\Day{\bullet}, \Day{i})$ is left and right residuated: there
          exists a function
          $\Day{\rightblackspoon} : \LowerSet{A} \times \LowerSet{A} \to
            \LowerSet{A}$ such that ... FIXME
    \item If $(\bullet, i)$ is a commutative pomonoid, then so is
          $(\Day{\bullet}, \Day{i})$. In this case, the residuals coincide.
  \end{enumerate}
  \bob{Could also say that $\eta$ preserves the monoids}
\end{proposition}

@@

\begin{proposition}\label{prop:lower-set-duoidal}
  If $(\bullet, i)$ and $(\lhd, j)$ are in a duoidal relationship then
  so are $(\Day{\bullet}, \Day{i})$ and $(\Day{\lhd}, \Day{j})$.
\end{proposition}

\subsubsection{Closed Lower Sets}

In the phase semantics, formulas are interpreted as elements that are
fixed points of a closure operator defined by a double negation with
respect to the monoid on the original frame. This closure operator
generates a partial order of ``facts'' whose meets and joins exactly
correspond to the syntactic ones when the original monoid is derived
from the proofs. As we mentioned in the introduction, the presence of
a self-dual operator in BV means that we cannot use double negation
closure, and we have to proceed more deliberately to preserve
join-like structure in an MAV-frame when building MAV-algebras.

We do this by defining closed lower sets as those that are closed
under finite $+$-combinations of their elements.

Lower sets add residuals to the pomonoids of an MAV-frame, so the Chu
construction below will allow us to add duals to interpret all of the
multiplicative subsystem of MAV. To incorporate the additives in such
a way that we will be able to derive the proof normalisation property,
we need to refine lower sets to ones that are closed under
$+$-combinations.

This corresponds to a sheaf construction, but we do not assume that
the ``join'' operation is stable with respect to meets.

The following development has proved amenable to mechanisation.

$+$-contexts are the semantic analogue of Horne's \emph{Killing
  contexts}.

\begin{definition}
  Let $(A, \leq)$ be a partial order with a monotone binary operation
  $+ : A \times A \to A$.  Let $F \in \LowerSet{A}$. A
  \emph{$+$-context} over $F$ is a binary tree built from constructors
  $\mathsf{leaf} : F \to \mathrm{ctxt}(F)$ and
  $\mathsf{node} : \mathrm{ctxt}(F) \times \mathrm{ctxt}(F) \to
    \mathrm{ctxt}(F)$. We define the \emph{sum} of a context as
  $\mathit{sum}(\mathsf{leaf}~x) = x$ and
  $\mathit{sum}(\mathsf{node}(c,d)) = \mathit{sum}(c) +
    \mathit{sum}(d)$.
\end{definition}

\begin{remark}
  Conceptually higher level view on contexts: if we view a lower set
  $F$ as a monotone function $F : A^\op \to \Omega$, then we can view
  the context construction as a relative monad on lower sets to
  partial orders.  \bob{and this should extend to presheaves and
    categories}.
\end{remark}

\begin{definition}
  A lower set $F \in \LowerSet{A}$ is \emph{$+$-closed} if for all
  $c \in \mathsf{ctxt}(F)$, $\mathit{sum}(c) \in F$. $+$-closed lower
  sets are ordered by subset inclusion and form a partial order
  $(\ClosedLowerSet{A}, \subseteq)$.
\end{definition}

\begin{proposition}
  There is a monotone function
  $\alpha : \LowerSet{A} \to \ClosedLowerSet{A}$ with the following
  properties. It is a Galois reflection. FIXME

  \begin{displaymath}
    \alpha(F) = \{ x \mid c \in \mathrm{ctxt}(F), x \leq \mathit{sum}(c) \}
  \end{displaymath}
\end{proposition}

\begin{proof}
  Explain the Agda, this relies on the fact that $\mathrm{ctxt}$ is a
  (relative) monad.
\end{proof}

@@

\begin{proof}
  $\alpha$ preserves the monoidal structure, so we can use it to prove
  the monoid laws for the constructed monoid.

  \bob{Sketch why this works: we can multiply together the contexts.}
\end{proof}

@@

\begin{proof}
  \bob{Explain the Agda here}
\end{proof}

@@

\begin{proposition}
  Let $(\bullet, i)$ and $(\lhd, j)$ be pomonoids on $(A, \leq)$ in a
  duoidal relationship, and assume that $(\bullet, i)$ distributes
  over $+$ (as in Proposition \ref{prop:closed-monoid-distrib}) and
  $(\lhd, j)$ is entropic with respect to $+$ (as in Proposition
  \ref{prop:closed-monoid-duoidal}). Then
  $(\ClosedDay{\bullet}, \ClosedDay{i})$ and
  $(\ClosedDay{\lhd}, \ClosedDay{j})$ are in a duoidal relationship on
  $(\ClosedLowerSet{A}, \subseteq)$.
\end{proposition}

\begin{proof}
  \bob{Explain the Agda here}
\end{proof}

\subsubsection{Chu Construction}

@@

We now turn to the self-dual duoidal structure required to interpret
the $\vSeq$ connective. First we transfer pomonoids from $(A, \leq)$
to self-dual pomonoids on $(\Chu(A, k), \sqsubseteq)$ provided they
interact well with $k$:
\begin{proposition}\label{prop:chu-self-dual}
  Let $(\lhd, j)$ be a pomonoid on $(A, \leq)$ such that
  $k \lhd k \leq k$ and $j \leq k$, then
  $x ; y = (x^+ \lhd y^+, x^- \lhd y^-)$ and $J = (j, j)$ form a
  self-dual pomonoid on $\Chu(A, k)$.
\end{proposition}

\begin{proof}
  \bob{show that that two extra bits are needed.}
\end{proof}

@@

\begin{proposition}
  If $(\bullet, i)$ and $(\lhd, j)$ are in a duoidal relationship on
  $(A, \leq)$, and $(\lhd, j)$ satisfies the conditions of Proposition
  \ref{prop:chu-self-dual}, then $(\otimes, I)$ and $(;, J)$ are in a
  duoidal relationship on $\Chu(A, k)$.
\end{proposition}

\begin{proof}
  \bob{Explain the Agda}
\end{proof}

>>>>

\section{Semantic Models for MAV}\label{sec:mav-semantics}

@@

\begin{itemize}
\item We construct algebraic models of MAV
\item MAV-algebras are algebras that MAV is sound for. In some
  sense, this is straightforward, but we identify precisely the
  structure that is needed.
\item The key structure is duoidality, or entropy.
\item We develop some model theory here, and show that MAV-algebras
  can be induced by the weaker structure of MAV-frames. This is
  essential for the cut-free elimination proof.
\end{itemize}

\subsection{Pomonoidal, $*$-autonomous, and Duoidal Structure on Partial Orders}

@@

\bob{Change $;$ to $\lhd$}

@@

\begin{remark}
  An MAV-frame is essentially two duoidal relationships and a
  distributivity law.
\end{remark}

\begin{remark}
  MAV-frames have a intuitive reading as a CCS-like process algebra
  (see Milner \cite{milner89} for an introduction to CCS). If we
  assume the existence of a collection of ``action'' elements
  $a \in F$ and their duals $\overline{a} \in F$, satisfying
  $a \parallel \overline{a} \leq I$, then we can read the constructs
  of an MAV-frame as parallel composition, sequential composition, and
  choice. The ordering is interpreted as a reduction relation. An
  interesting avenue for future work would be to discover to what
  extent MAV can be thought of as a logic for processes in this
  process algebra.
\end{remark}

@@

We construct MAV-algebras from MAV-frames in a three step process. In
Section \ref{sec:lower-sets}, we use lower sets and the Day
construction to add meets, joins and residuals for pomonoids to a
partial order. This construction creates joins freely, so we restrict
to $+$-closed lower sets in Section \ref{sec:closed-lower-sets}, which
separates the Day construction of pomonoids into two separate
cases. Finally, we create the $*$-autonomous structure using the Chu
construction in Section \ref{sec:chu}. We maintain the duoidal
structure through each construction.

\subsubsection{Lower Sets and Day pomonoids}
\label{sec:lower-sets}

@@

\begin{proposition}
  For any $(A, \leq)$, the function $\eta$ is monotone, and
  $(\LowerSet{A}, \subseteq)$ has meets and joins given by
  intersection and union respectively.
%
%  \bob{and $\eta$ preserves any meets that exist}
\end{proposition}

\begin{proposition}\label{prop:day-construction}
  If $(\bullet, i)$ is a pomonoid on $(A, \leq)$, then there is a
  corresponding \emph{Day pomonoid} $(\Day{\bullet}, \Day{i})$ on
  $\LowerSet{A}$ defined as
  $F \Day\bullet G = \{ z \mid z \leq x \bullet y, x \in F, y \in G
    \}$ and $\Day{i} = \eta(i)$. Moreover:
  \begin{enumerate}
    \item $(\Day{\bullet}, \Day{i})$ is left and right residuated: there
          exists a function
          $\Day{\rightblackspoon} : \LowerSet{A} \times \LowerSet{A} \to
            \LowerSet{A}$ such that ... FIXME
    \item If $(\bullet, i)$ is a commutative pomonoid, then so is
          $(\Day{\bullet}, \Day{i})$. In this case, the residuals coincide.
  \end{enumerate}
  \bob{Could also say that $\eta$ preserves the monoids. Does this
    come up later?}
\end{proposition}

@@

\begin{proposition}\label{prop:lower-set-duoidal}
  If $(\bullet, i)$ is duoidal over $(\lhd, j)$ then
  $(\Day{\bullet}, \Day{i})$ is duoidal over $(\Day{\lhd}, \Day{j})$.
\end{proposition}

\subsubsection{$+$-closed Lower Sets}
\label{sec:closed-lower-sets}

In the phase semantics, formulas are interpreted as elements that are
fixed points of a closure operator defined by a double negation with
respect to the monoid on the original frame. This closure operator
generates a partial order of ``facts'' whose meets and joins exactly
correspond to the syntactic ones when the original monoid is derived
from the proofs. As we mentioned in the introduction, the presence of
a self-dual operator in BV means that we cannot use double negation
closure, and we have to proceed more deliberately to preserve
join-like structure in an MAV-frame when building MAV-algebras. We do
this by defining $+$-closed lower sets as those that are closed under
finite $+$-combinations of their elements. This leads to a closure
operator on lower sets that allows us to immediately deduce that
$+$-closed lower sets form a lattice. We also preserve the
Day-pomonoids from lower sets, but in two different ways, depending on
how the original pomonoid interacts with $+$. In Proposition
\ref{prop:closed-monoid-distrib} we handle pomonoids that distribute
over $+$, and in Propositional \ref{prop:closed-monoid-duoidal} we
handle pomonoids that are entopic under $+$. We need these
constructions to lift the $\parallel$ and $\lhd$ pomonoids from
MAV-frames to $+$-closed lower sets. Finally in this section, we show
that duoidal structure on lower sets from Proposition
\ref{prop:lower-set-duoidal} is preserved in $+$-closed lower sets.

For this section, we assume that $(A, \leq)$ is a partial order with a
monotone binary operation $+ : A \times A \to A$ (we do not assume
that $+$ is a join or even a pomonoid.)

\begin{definition}
  A lower set $F \in \LowerSet{A}$ is \emph{$+$-closed} if $x \in F$
  and $y \in F$ imply $x + y \in F$. $+$-closed lower sets are ordered
  by subset inclusion and form a partial order
  $(\ClosedLowerSet{A}, \subseteq)$.
\end{definition}

\begin{proposition}
  Let $U : \ClosedLowerSet{A} \to \LowerSet{A}$ be the ``forgetful''
  function that forgets the $+$-closed property. There is a monotone
  function $\alpha : \LowerSet{A} \to \ClosedLowerSet{A}$ such that
  for all $F \in \ClosedLowerSet{A}$, $\alpha(U F) = F$ and for all
  $F \in \LowerSet{A}$, $F \subseteq U (\alpha F)$.
\end{proposition}

\begin{proof}
  To define $\alpha$, we close lower sets under all
  $+$-combinations. To this end, for $F \in \LowerSet{A}$, define
  $\mathrm{ctxt}(F)$, the set of all $+$-combinations of $F$
  inductively built from constructors
  $\mathsf{leaf} : F \to \mathrm{ctxt}(F)$ and
  $\mathsf{node} : \mathrm{ctxt}(F) \times \mathrm{ctxt}(F) \to
  \mathrm{ctxt}(F)$. We define the \emph{sum} of a context as
  $\mathit{sum}(\mathsf{leaf}~x) = x$ and
  $\mathit{sum}(\mathsf{node}(c,d)) = \mathit{sum}(c) +
  \mathit{sum}(d)$. Now define:
  $\alpha(F) = \{ x \mid c \in \mathrm{ctxt}(F), x \leq
  \mathit{sum}(c) \}$. This is $+$-closed, by taking the
  $\mathsf{node}$ combination of contexts. $\alpha \circ U$ is
  idempotent because $\alpha$ does not introduce any elements to lower
  sets that are already closed. For arbitrary lower sets $F$,
  $F \subseteq U(\alpha F)$ by the $\mathsf{leaf}$ constructor.
\end{proof}

@@

\begin{proof}
  We can define an operation
  $\bullet^c : \mathrm{ctxt}(F) \times \mathrm{ctxt}(G) \to
  \mathrm{ctxt}(F \Day{\bullet} G)$ that ``multiplies'' two trees,
  such that
  $\mathit{sum}(c) \bullet \mathit{sum}(d) \leq \mathit{sum}(c
  \bullet^c d)$, using the distributivity. This allows us to show that
  $\alpha$ preserves the monoid operation:
  $\alpha F \ClosedDay{\bullet} \alpha F = \alpha (F \Day{\bullet}
  G)$. With this, we can show that the monotonicity, associativity,
  unit, and commutativity properties of $\Day{\bullet}$ transfer over
  to $\ClosedDay{\bullet}$. The definition of the residual from lower
  sets is already $+$-closed, by distributivity.
\end{proof}

@@

\begin{proof}
  Since $+$ is entropic over $(\lhd, j)$, the Day monoid $\Day{\lhd}$
  is automatically $+$-closed. The monoid structure directly
  transfers. Similarly, $\eta(j)$ is automatically $+$-closed since
  $j + j \leq j$.
\end{proof}

@@

\begin{proposition}
  Let $(\bullet, i)$ and $(\lhd, j)$ be pomonoids on $(A, \leq)$ in a
  duoidal relationship, and assume that $(\bullet, i)$ distributes
  over $+$ (as in Proposition \ref{prop:closed-monoid-distrib}) and
  $+$ is entropic over $(\lhd, j)$ (as in Proposition
  \ref{prop:closed-monoid-duoidal}). Then
  $(\ClosedDay{\bullet}, \ClosedDay{i})$ and
  $(\ClosedDay{\lhd}, \ClosedDay{j})$ are in a duoidal relationship on
  $(\ClosedLowerSet{A}, \subseteq)$.
\end{proposition}

\begin{proof}
  The entropic relationship established in Proposition
  \ref{prop:lower-set-duoidal} carries over thanks to the properties
  of $\alpha$ and $U$. The fact that
  $\ClosedDay{i} \subseteq \ClosedDay{j}$ relies on the condition
  $j + j \leq j$ to collapse $+$-contexts of $j$s.
\end{proof}

\subsubsection{Chu Construction}
\label{sec:chu}

@@

We now turn to the self-dual duoidal structure required to interpret
the $\vSeq$ connective. First we transfer pomonoids from $(A, \leq)$
to self-dual pomonoids on $(\Chu(A, k), \sqsubseteq)$ provided they
interact well with $k$:
\begin{proposition}\label{prop:chu-self-dual}
  Let $(\lhd, j)$ be a pomonoid on $(A, \leq)$ such that $(\bullet, i)$ is entropic over 
  $k \lhd k \leq k$ and $j \leq k$, then
  $x ; y = (x^+ \lhd y^+, x^- \lhd y^-)$ and $J = (j, j)$ form a
  self-dual pomonoid on $\Chu(A, k)$.
\end{proposition}

\begin{proof}
  $x;y$ is well-defined because
  $(x^+ \lhd y^+) \bullet (x^- \lhd y^-) \leq (x^+ \bullet x^-) \lhd
  (y^+ \bullet y^-) \leq k \lhd k \leq k$. $J$ is well defined because
  $j \bullet j \leq j \leq k$. The pomonoid laws all transfer directly.
\end{proof}

@@

\begin{proposition}
  If $(\bullet, i)$ is entropic over $(\lhd, j)$ on $(A, \leq)$, and
  $(\lhd, j)$ satisfies the conditions of Proposition
  \ref{prop:chu-self-dual}, then $(\otimes, I)$ and $(;, J)$ are in a
  duoidal relationship on $\Chu(A, k)$.
\end{proposition}

\begin{proof}
  For the positive half of the Chu construction, this is a direct
  consequence of the entropy on $(A, \leq)$. For the negative half, we
  use Lemma \ref{lem:duoidal-residual} and the fact that meets are
  always duoidal.
\end{proof}
